### 2016-03-20 doc/draft.Rmd
discussion outline

# plot of read start postion vs. genome
# do we not find the same variants
```

@@

We have shown that benchmarking is vital to accurate NGS based studies of viral diversity. In our opinion DeepSNV represents the most rigorously benchmarked SNV indentifier. We show that even under the best conditions the accuracy of DeepSNV drops when applied to whole genome sequencing of an RNA virus. The accuracy of SNV identification under more realistic conditions was even more sobering. We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and thatat moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions. We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates. While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in acurate SNV detection and few variant callers have been benchmarked under patient derived conditions . Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying this error.

 To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS. Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus. However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8. The most stiking drop in sensitivity came for variants found at 0.16%. Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported 0.860 for variants at 0.1%. This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene. Therefore, our sequencing efforts are diluted and we expect fewer reads per variant which limits sensitivity. Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants and this would play a larger role for rare variants that have fewer reads to begin with. Dispite the use of bonferronni corrected p values, the relatively modest decrease in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).

Because, our WSN3-PR8 populations were constructed from reconstituded cDNA genomes they were devoid of any biases that might accompany PCR amplification. Gerstung *et. al.* have shown that PCR amplification biases lead to overdispered data. We found that the dispersion parameter estimated by DeepSNV was relatively low (<5) indicating that the variability in the data was relatively close to what would be expected in a binomial distribution. On the other hand, the dispersions estimated for our 20 mutant populations were much higher (>100) indicating more variability in the data. This variability is mostlikely due to the RT-PCR based biases as has been observed in the past, and likely accounts for the drastic drop in specificity that accompanied our 20 mutant data set. We were able to increase specificity through a number of empirically determined quality cut offs. We have sequenced over 300 influenza libraries in on more that 5 Hiseq lanes and the mean MapQ and mean Phred distributions for subconsensus variants is consistant. We trust then that our cut offs are valid when applied to influenza sequencing. However, these distributions are likely influenced by the articheture of the influenza geneome which contains conserved ends of every fragment. We expect poorly mapping reads to map to these conserved areas and so by removing this reads we are indirectly removing variants that are being called in the poorly covered regions of the genome.

One of the take home points of this manuscript is that accuracy is a function of the conditions under which a test is applied. Therefore to determine the accuracy of DeepSNV in indentifying SNV from patient derived influenza populations we

>>>>

# plot of read start postion vs. genome
# do we not find the same variants
# L1-norm of 4_5 and 5_5
```

@@

We have shown that benchmarking is vital to accurate NGS based studies of viral diversity. In our opinion DeepSNV represents the most rigorously benchmarked SNV indentifier. We show that even under the best conditions the accuracy of DeepSNV drops when applied to whole genome sequencing of an RNA virus. The accuracy of SNV identification under more realistic conditions was even more sobering. We found input nucleic acid concentrations, which can vary greatly in patient-derived samples, had large effects on both sensitivity and specificity, and that at moderate levels of nucleaic acid input we could greatly improve accuracy by filtering putative SNV based on quality metrics and read positions. We further improved our accuracy at low input levels by controlling for RT-PCR errors by processing samples in duplicates. While our quality cut offs may not be needed by all variant callers, our work with Lofreq indicates that experimental design is important in acurate SNV detection and few variant callers have been benchmarked under patient derived conditions . Finally, we show that these inaccuracies in SNV calling drastically impact down stream analysis, and that accurate conclusions require quanitifying and limitting this error.

 To our knowledge, DeepSNV is the only variant calling method that has been benchmarked on real sequencing reads in which all true positive variants and their frequencies were known *a priori* and independent of NGS. Given this robust benchmarking we thought it ideal for identifying rare variants in populations of influenza virus. However, we found decreased sensitivity and specificity when applied to cell culture derived mixtures of WSN33 and PR8. The most stiking drop in sensitivity came for variants found at 0.16%. Under these conditions we found true variants at a rate of `r miseq.sense.16` compared to the reported sensitivity of  0.860 for variants at 0.1%. This decrease is most likely due to the fact that we are looking for variants across `r miseq_kb`bp of the influenza genome where previous benchmarking focused on just 1,500 bp of the HIV pol gene. Therefore, our sequencing efforts are diluted and we expect fewer reads per variant which limits sensitivity. Additionally, in our analysis we aligned all reads to a PR8 reference genome which biases our detection of WSN33 variants. Any bias against WSN33 would decrease our detection of rare variants, which already have few reads. A relatively modest decrease in specificity accompanied the reduced sensitivity.Dispite the use of bonferronni corrected p values, the drop in specificity is mostlikely due to multiple testing errors caused by applying the algorithm across `r miseq_kb`bp (cite).

 Despite this reduced accuracy we applied DeepSNV to experimental populations that more closely mimic the diversity and nucleaic acid input levels found in patient derived samples. At a modest input level of 10^5^ genomes per microliter we found roughly a four-fold increase in false positives. This decrease is likely dues to how the populations were constructed and stresses the importance of benchmarking SNV identifying pipelines under conditions to which they will be applied.
III. Experimental populations
  A. The large increase in false positives compared to PR8-WSN33
    i) Our WSN3-PR8 populations were constructed from reconstituded cDNA genomes and were devoid of any biases that might accompany PCR amplification. 
    ii) Not the case with the experimental samples
    iii) Gerstung *et. al.* have shown that PCR amplification biases lead to increased variability in the subconsensus nucleotide counts used to call SNV. This variability in erreoneous base calls in test samples compared to the control increases the likelihood \ that the frequency of a false positive SNV will exceed the error rate estimated by the plasmid control.
    iv) DeepSNV can attempt to control for such variablity inpart by estimating the dispersion of a sample. Indeed, our experimental samples were more dispersed than those in cell culture. However, the variability estimates are limitted  and this likely accounts for the decreased specificity.
  B. Quality cut offs
    We hypothesis that the RT-PCR biases introduced in our experimental populations reduced the efficacy of teh DeepSNV algorithm. We,therefore, used alternative means of increasing accuracy.
    i) The average MapQ and Phred scores of putative SNV were bimodal with true SNV lying in the higher of the two distributions. 
    ii) We plotted the cdf of these distributions and applied cuttoffs at the streepest increase in the true positive distributions.
    iii) We have sequenced over 300 influenza libraries on over 5 Hiseq lanes and observe the consistant qualty distributions for DeepSNV identified putative SNV.
    iv) We have also sequenced many polio libraries and see the same bimodal trends; with slightly shift means.
    v) Mostlikely due to influenza genomic structures.
    vi) This is different than filtering reads and masking phred, as we are filtering SNV call after they have been identified. Treats False positives and True negatives differently, and perhaps inflates coverage estimates.
    vi) MapQ, Phred, and Read position were not correlated with eachother and so remove different false positive variants
    vii) We found many false SNV were identified only at the termi of paired end reads. We removed these by filtering putatuive SNV calls based on their position on a paired end read; however, there are many ways to account for these errors. 
    viii) Strand bias, trimming
    ix) In particular these were found almost exclusively in regions of the genome that were enriched for read start sites. This enrichment is mostlikely a consequence  of sequence context, the fragmentation process, and our size selection protocol; however, it is unclear why our plasmid control would not contain similar biases. The cause may have a biological foundation in DI particals which have been identified to conatian truncated genomic segements and would only be present in infected supernantants and not a plasmid control.
    x) These ammendments were able to greatly increase the accuracy of our DeepSNV based SNV pipeline.

  C. Discussion of frequency of variants
    i) due to variability in frequency measurments cut offs limited sensitivity we did not apply frequency cut offs as is commonly done
    ii) Before our cut off no variants were found below XXX%. after our cut offs non were found below X% showing we indirectly applied a frequency threshold.
    iii) discussion of coverage
    iv) Lofreq and frequency cut offs.
    v) reiterate importance of decreases sensitivity when comparing rare SNV between samples.

  D. Lower inputs
  i) RT-PCR errors resulted in decreased specificity at lower levels of nucleic acid concentration input. 
  ii) As these errors result from base changes, they appear as TP to algorithms and have high quality scores.
  iii) Where contorled for by processing samples in duplicate.
  iv) could duplicates have been pooled and sequenced as one sample?

IV. Comparision to Lofreq  and Consequences of errors
  i) our benchmarking shows that DeepSNVs means of distinguishing true from false variants is partitially breaks down under conditions that mimic patient derived samples.
  ii) This can be overcome with the addition of specific quality filters.
  iii) We would not have known about these inaccuracies had we not run our benchmarking experiment.
  iv) Do other variant callers have similar inaccuracies.
  v) Yes Lofreq, although more accurate than our unmodified DeepSNV pipeline, had decreased specificity than had previously been reported.
  vi) The initial Lofreq sensitivity and specificity metrics were based largely on a *in silico* data sets.
  vii) Lofreq did not display this high accuracy in under our experimental conditions.
  viii) While we did not robustly attempt to improve the Lofreq accuracy we did find that frequency threshold could improve its specificity.
  ix) Experimental conditions matter, regardless of variant calling method and so accurate study requires controls to be run.

V. Consequences of errors
  A. 
  i) Accuracy in SNV indetification has consequences on down stream analysis.
  ii) Diversity measurements and populations comparisions are affected.
  iii) Differences in titer must be accounted for in population comparisions.

VI. Take home message
i) Benchmarking controls are needed to ensure accurate NGS based studies of viral populations.
i) As influenza virologist we benefit from an easily accessible reverse genetics system that made for a straight forward way to benchmark our variant calling.
ii) Hopefully we have convince the reader that such controls are necessisary.
iii) While we do not recomemd a varaints calling pipeline we have provided our raw data and computational approach.
iv) It is provides the most robust data set to data and avaialable for anyone to test their methods against.
